---
title: "Pricing defaultable bonds in the shadow-intensity framework"
author: "Jean-Paul Renne"
date: "HEC Lausanne"
bibliography: ["ref.bib"]
biblio-style: "apalike"
link-citations: true
output:
  #bookdown::pdf_document2: default
  bookdown::html_document2:
    toc: true
    toc_float: true
  number_sections: true
---



[//]: <> (Below: change in title format:)

<style type="text/css">
body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 30px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 24px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 18px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 14px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# The context 

@Pallara_Renne_2023 and @Renne_Pallara_2023 investigate the pricing the sovereign credit risk in the context of shadow-intensity frameworks. In that type of framework, the prices of defaultable bonds are given by conditional expectations of the form:
\begin{equation}
\mathbb{E}_t(\exp[\color{blue}{-z_{t+1}-\dots-z_{t+h}}-\max(0,\lambda_{t+1})-\dots -\max(0,\lambda_{t+h})]),(\#eq:condExp)
\end{equation}
where $z_t$ (risk-free short-term rate) and $\lambda_t$ (*shadow default intensity*) are affine functions of the state vector $x_t$. In addition, $x_t$ follows a Gaussian vector auto-regressive (VAR) model:
\begin{eqnarray*}
z_t &=& \dot{a}'x_t + \dot{b}\\
\lambda_t &=& a'x_t + b,
\end{eqnarray*}
and
\begin{equation}
x_t = \mu + \Phi x_{t-1}+ \Sigma\varepsilon_t,(\#eq:var)
\end{equation}
with $\varepsilon_t \sim \mathcal{N}(0,Id)$ (where $Id$ denotes the identity matrix).

Equation \@ref(eq:condExp) gives the risk-neutral price of a bond of maturity $h$, issued by an entity whose default intensity is $\underline{\lambda}_t = \max(0,\lambda_t)$. (That is, the conditional default probability of the issuer, on date $t$, is $\exp(-\underline{\lambda}_t)$.)

Absent the blue terms, the conditional expectations appearing in \@ref(eq:condExp) would be of the same type as the ones considered in the shadow-rate literature (see @Black_1995). In that case, one could for instance use the approaches proposed by @Priebsch_2013 or @Wu_Xia_2016 to calculate approximations for these conditional expectations. @Pallara_Renne_2023 and @Renne_Pallara_2023 extend the approach of @Wu_Xia_2016 to compute approximate values to \@ref(eq:condExp). Th present page uses the latter approach, employing codes contained in package `TSModels` to implement this approach.

To start with, load package `TSModels`. This library is available on GitHub and can be accessed using the following lines of code:

```{r essai, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("jrenne/TSModels")
```


# A numerical example

## Specifications

Consider a two-factor model, that is $x_t = [x_{1,t},x_{2,t}]'$. Assume that $z_t$ (that can be interpreted as the risk-free short term rate) depends on $x_{1,t}$ only, and that $\lambda_t$ (the shadow default intensity) depends on $x_{2,t}$.

We consider the following VAR dynamics for $x_t$:
$$
x_t = \left[\begin{array}{cc}0.9 & 0 \\ 0 & 0.9\end{array}\right] x_{t-1} +
\left[\begin{array}{cc}1 & 0.1 \\ 0 & 1\end{array}\right]\varepsilon_t,
$$
which determines $\mu$, $\Phi$, and $\Sigma$ of \@ref(eq:var).

Further, we want $z_t$ (respectively $\lambda_t$) to be of mean 2\% (respectively 0) and of standard deviation $1\%$ (resp. 2/%).

Let us code these specifications:

```{r settings}
Phi_x <- .9 * diag(2) # set Phi
Sigma_x <- diag(2) # set Sigma
Sigma_x[1,2] <- .1
Mu_x <- matrix(0,2,1) # set mu

stdv.z      <- .01 # set the unconditional std dev of z
stdv.lambda <- .02 # set the unconditional std dev of lambda

# Compute the unconditional variance of x_t:
uncond.Variance.X <- matrix(solve(diag(2*2) - Phi_x %x% Phi_x) %*%
                              c(Sigma_x %*% t(Sigma_x)),2,2)

# Deduce a and a.dot:
a.dot <- matrix(c(stdv.z/sqrt(uncond.Variance.X[1,1]),0))
a     <- matrix(c(0,stdv.lambda/sqrt(uncond.Variance.X[2,2])))

# Set the unconditional means of and lambda:
b.dot <- .02
b     <- .0
```

## Simulations

Let us have a look at simulated paths for $z_t$ and $\lambda_t$:

```{r simulation1}
library(TSModels)

nb.periods <- 200
X <- simul.X(Phi_x,Mu_x,Sigma_x,nb.periods)
z <- b.dot + a.dot[1] * X[1,1,]
lambda <- b + a[2] * X[2,1,]
```

```{r simulation2, fig.cap="This plots simulated paths for $z_t$ (risk-free short-term rate) and $\\lambda_t$ (shadow intensity). The blue dotted line, on the right-hand-side plot, is the effective intensity, that is $\\max(0,\\lambda_t)$. The vertical lines indicate those dates where lambda is minimal and maximal.", echo=FALSE}
par(mfrow=c(1,2))
plot(z,type="l",xlab="",main=expression(z[t]))
plot(lambda,type="l",xlab="",main=expression(lambda[t]))
abline(h=0,col="grey")
lines(pmax(0,lambda),col="blue",lwd=2,lty=3)
t.min.lambda <- which(lambda==min(lambda))
t.max.lambda <- which(lambda==max(lambda))
abline(v=t.min.lambda,col="red")
abline(v=t.max.lambda,col="red",lty=2)
```

# Bond pricing {#sec:Model}

Let us now evaluate the conditional expectations of \@ref(eq:condExp), for $h=\{1,\dots,10\}$. For that, we make use of function `compute.condit.Exp` of package `TSModels`.

```{r computeExp, fig.cap="This is the caption"}
H <- 10
res <- compute.condit.Exp(a,b,a.dot,b.dot,
                          Phi_x,Mu_x,Sigma_x,
                          X=cbind(X[1,1,],X[2,1,]),
                          max.h = H)
t.min.lambda <- which(lambda==min(lambda))
t.max.lambda <- which(lambda==max(lambda))
rbind(res$E.n[t.max.lambda,], #approximate values of the conditional expect. for date t
      -log(res$E.n[t.max.lambda,])/(1:H))
```

The last two lines respectively give, for the date featuring the highest $\lambda_t$, the term structures of the bond prices and of associated yields-to-maturity.

Let us now compute credit-risk-free bond prices. This is obtained by taking $a=0$ and $b=0$ (so that $\lambda_t=0$): 

```{r riskfree, fig.cap="This is the caption"}
res0 <- compute.condit.Exp(a*0.00001,b*0.00001,a.dot,b.dot,
                          Phi_x,Mu_x,Sigma_x,
                          X=cbind(X[1,1,],X[2,1,]),
                          max.h = H)
y       <- - log(res$E.n) / t(matrix(1:H,H,nb.periods))
y_RF    <- - log(res0$E.n) / t(matrix(1:H,H,nb.periods))
spreads <- y - y_RF
```

Let us plot the term strcutures of yields and spreads for the date on which we observe the highest $\lambda_t$:

```{r chartspreads, fig.cap="Term structures of risk-free yields (in red), and credit-risky yields (in black). Solid lines (respectively dashed lines) correspond to the date where $\\lambda_t$ is the highest (resp. the lowest). These two dates are indicated by vertical lines on the previous figure."}
plot(y[t.max.lambda,],type="l",ylim=c(0,max(y)),xlab="maturity",ylab="yields",
     main="Term structures of bond yields (risk-free and defaultable bonds)",lwd=2)
lines(y_RF[t.max.lambda,],col="red",lwd=2)
lines(y_RF[t.min.lambda,],col="red",lwd=2,lty=2)
lines(y[t.min.lambda,],lwd=2,lty=2)
```


# Check formula using simulations

In this section, we calculate the conditional expectation given in \@ref(eq:condExp) using Monte-Carlo simulations, for different maturities $h$. We will then compare these prices with the ones resulting from function `compute.condit.Exp`. 

We consider the date on which $\lambda_t$ is the highest.


```{r loaddataset, fig.cap="This is the caption"}
nb.replics <- 10000 # Number of simulated trajectories
t  <- t.max.lambda # date on which cond. expect. are computed
Xt <- matrix(X[,1,t],ncol=1) # state vector on that date

# Simulate trajectories starting from Xt:
X <- simul.X(Phi_x,Mu_x,Sigma_x,nb.periods=H,X0 = Xt,nb.replics = nb.replics)
# Compute prices for each trajectory:
aux1 <- apply(X,c(2,3),function(x){exp(- b.dot - t(x)%*%a.dot -
                                        pmax(0,b + t(x)%*%a))})
aux2 <- apply(aux1,1,cumprod)
# Compute averages of prices across simulated trajectories:
simul.prices <- apply(aux2,1,mean)
# Compare with output of compute.condit.Exp:
plot(res$E.n[t,],lwd=2,main="Approximate formula versus Monte-Carlo approach")
lines(simul.prices)
legend("topright",c("Approximate formula","Monte-Carlo"),lty=c(NaN,1),pch=c(1,NaN),lwd=c(2,1))
```

# References {-}

